\section{Conclusions}
\label{sec:conclusions}

We identified three challenges in LID estimation and explained how the existing methods do satisfy those desiderata. 
To overcome those limitations we introduced an algorithm for LID estimation which relies on powerful neural parametric density estimators, and provided solid theoretical justification for the method. 
Our experiments showed that it can scale to datasets of thousands of dimensions and give accurate estimates on complicated manifolds. 
We investigated its strengths and limitations and showed that LID is connected with local model performance, 
especially in unsupervised learning and classification settings.


There is a number of future research directions stemming from this work. 
The first one is a more theoretically grounded and experimentally tested procedure for choosing $\delta$ in the presence of observation noise, which might be important for practitioners. 
Another one is further investigating the connection of LID estimates and classifier performance: LID estimates could be used in active learning, semi supervised learning or curriculum learning. 

\section{Acknowledgements}
Some experiments were performed using the Entropy cluster at the Institute of Informatics, University of Warsaw, funded by NVIDIA, Intel, the Polish National Science Center grant UMO2017/26/E/ST6/00622 and ERC Starting Grant TOTAL. 
Some experiments were performed using the GUŚLARZ 9000 workstation at the Polish National Institute for Machine Learning.

Jacek Tabor is supported by Foundation for Polish Science (grant no POIR.04.04.00-00-14DE/18-00) carried out within the Team-Net program co-financed by the European Union under the European Regional Development Fund.
Przemysław Spurek is supported by the National Centre of Science (Poland) Grant No. 2019/33/B/ST6/00894.
Adam Goliński is supported by the UK EPSRC CDT in Autonomous Intelligent Machines and Systems.

We wish to thank various people for their contribution to this project: Marek Cygan for overall support during this project and valuable and constructive suggestions on the manuscript; Maciej Dziubiński, Piotr Kozakowski, Dominik Filipak, and Maciej Śliwowski for their valuable and constructive suggestions on the manuscript.

\section{CRediT Author Statement}

\begin{table}[H]
\renewcommand{\arraystretch}{1.4}

\resizebox{\columnwidth}{!}{%
\begin{tabular}{lllllll}
 & \rotatebox{70}{PT} & \rotatebox{70}{RM} & \rotatebox{70}{ŁG} & \rotatebox{70}{PS} & \rotatebox{70}{JT} & \rotatebox{70}{AG} \\ \cline{2-7} 
 
\multicolumn{1}{l|}{Conceptualization} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} \\ \cline{2-7} 
\multicolumn{1}{l|}{Methodology} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}{\color[HTML]{656565} }} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} \\ \cline{2-7} 
\multicolumn{1}{l|}{Software} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}{\color[HTML]{343434} }} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}{\color[HTML]{656565} }} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} \\ \cline{2-7} 
\multicolumn{1}{l|}{Validation} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} \\ \cline{2-7} 
\multicolumn{1}{l|}{Formal analysis} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} \\ \cline{2-7} 
\multicolumn{1}{l|}{Investigation} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} \\ \cline{2-7} 
\multicolumn{1}{l|}{Writing - Original Draft} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} \\ \cline{2-7} 
\multicolumn{1}{l|}{Writing - Review \& Editing} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}{\color[HTML]{656565} }} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} \\ \cline{2-7} 
\multicolumn{1}{l|}{Visualization} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} \\ \cline{2-7} 
\multicolumn{1}{l|}{Supervision} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} \\ \cline{2-7} 
\multicolumn{1}{l|}{Project administration} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} \\ \cline{2-7} 
\multicolumn{1}{l|}{Data Curation} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{\cellcolor[HTML]{343434}{\color[HTML]{9A0000} }} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} \\ \cline{2-7} 
\end{tabular}%
}
\end{table}
 
 


\comment {
\textbf{Piotr Tempczyk}: Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Writing - Original Draft, Writing - Review \& Editing, Visualization, Supervision, Project administration.
\textbf{Rafał Michaluk}: Software, Investigation, Data Curation.
\textbf{Łukasz Garncarek}: Methodology, Software, Formal analysis, Writing - Original Draft, Writing - Review \& Editing.
\textbf{Przemysław Spurek}: Writing - Original Draft, Writing - Review \& Editing.
\textbf{Jacek Tabor}: Conceptualization, Methodology, Writing - Review \& Editing, Supervision.
\textbf{Adam Goliński}: Conceptualization, Methodology, Writing - Original Draft, Writing - Review \& Editing, Visualization, Supervision.
}